\graphicspath{{chapters/papers/06/images}}
\chapter{TPES: timor purity estimation from SNVs}\label{ch:tpes}

\section{Abstract}
Tumour purity is the proportion of cancer cells in a tumour sample.
It impacts on the accurate assessment of molecular and genomics features.

	\subsection{Introduction}
	Genomic and molecular analysis of tumour samples require the quantification of tumour and admixed normal cells proportion.
	In order to asses the somatic lesion detection boundaries and to perform comparative analyses several tools were built to quantify TP from NGS data.
	The approaches based on SCNA fall short for samples with quiet genomes.
	To solve this purity can be estimated through the distribution of variant allelic fractions within copy number neutral tumour segments.

\section{Materials and methods}
The VAF distribution of a set of clonal monoallelic SNVs from a pure tumour sample should be centred in $0.5$.
Technical and cancer specific factors may influence the VAF value as reference mapping bias.
Moreover in the case of subclonal events the VAF is altered.
Clonal monoallelic SNVs in a diploid segment are suited for TP estimation and are named p-SNV.
Given a set of p-SNVs, TP could be computed as:

$$\frac{observerd\ VAF(pSNP)}{expected\ VAF}$$

Where $obvserved\ VAF$ is computed from the tumour data while $expected\ VAF$ is the value expected from a pure tumour sample accounting for reference mapping bias.
p-SNVs are selected with a conservative procedure.
To minimize the number of false positive p-SNVs for each sample, TPES introduces two main filtering steps.
In the first SNVs are selected from copy-number neutral segments applying a conservative filter on the Log R value of each genomic segment.
Moreover the log R is adjusted for ploidy and SNV are retained only with a number of reads mapping the alternative base and AG above and below threshold.
Chromosome $X$ and $Y$ are excluded to avoid gender stratification.
This nominates a set of heterzygous copy-number neutral SNPv.
The second filter TPES removes putative subclonal mutations.
Observed VAF distribution is smoothed by kernel density estimation.
Local maxima of the underlying distribution can be observed.
The peak with the highest VAF value is the candidate observed VAF.
